---
title: "Copy Cat: A recreation of McDonald et al's stable isotope dietary analysis of cats"
author: "John Cornelison"
date: "2025-06-24"
output:
  html_document:
    code_folding: show  # or "hide"
    toc: true           # optional: adds a table of contents
    toc_depth: 2 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo = FALSE, warning = FALSE, message = FALSE}
#Libraries
library(simmr)
library(MixSIAR)
library(tidyverse)
library(plyr)
library(DT)

raw_cat_data <- read_csv("raw_cat_data.csv")
raw_food_data <- read_csv("raw_food_data.csv")
raw_prey_data <- read_csv("raw_prey_data.csv") %>% 
  mutate(d13C = as.numeric(d13C),
         d15N = as.numeric(d15N))
posterior_probabilities <- read_csv("McDonalPaperPosteriorProbabilities.csv") %>%
  fill(Cat, .direction = "down") %>% 
  fill(Identifier, .direction = "down") %>% 
  mutate(variable = case_when(variable == ".globabl.CombinedPrey" ~ "p.global.CombinedPrey",
                              TRUE ~ variable),
         McDonald_Classification = case_when(`50%` >= 0.245 ~ "Hunter",
                                             `50%` < 0.245 ~ "Non Hunter",
                                             TRUE ~ "Unkown"),
         `CI_width` = `97.50%`-`2.50%`) %>% 
  filter(Variable == "p.global.CombinedPrey")
  filter()


```



### McDonald Et al paper

```{r, echo = FALSE, message = FALSE, warning = FALSE}


# 1. Read in the raw CSVs ---------------------------------------
prey_raw <- read_csv("raw_prey_data.csv") 
food_raw <- read_csv("raw_food_data.csv")
cat_raw  <- read_csv("raw_cat_data.csv")

# Filter to only samples marked "Y" for analysis
prey  <- prey_raw  %>% filter(`Part of Analysis?` == "Y")
food  <- food_raw  %>% filter(`Part of Analysis?` == "Y")
cats  <- cat_raw   %>% filter(`Part of Analysis?` == "Y") 

# Restrict to Southeast US cats (the only ones analyzed in the paper)
cats_se <- cats %>% filter(`Region` == "Southeast")

# 2. Define trophic‐enrichment (TEF) -----------------------------
# As per McDonald et al., Δ13C = +2.6‰, Δ15N = +1.9‰
TEF_means <- c(2.6, 1.9)
TEF_sds   <- c(0.10, 0.10)  # small uncertainty

# 3. Build the “CombinedPrey” source -----------------------------
prey_means <- prey %>% 
  summarise(d13C = mean(d13C), d15N = mean(d15N)) %>% 
  as.numeric()

prey_sds <- prey %>% 
  summarise(d13C = sd(d13C),   d15N = sd(d15N))   %>% 
  as.numeric()

# 4. Aggregate food by Brand + Type --------------------------------
# First tag whether each food sample is dry vs wet
food2 <- food %>% 
  mutate(
    type = if_else(is.na(`$/can (WF)`), "Dry", "Wet"),
    # further split wet into fish‐based vs meat‐based (simple keyword match)
    type = case_when(
      type == "Wet" & str_detect(Flavor, regex("tuna|salmon|ocean|fish", ignore_case=TRUE)) ~ "WetFish",
      TRUE ~ type
    )
  )

# Then average replicates by Brand & type
food_src <- food2 %>% 
  group_by(Brand, type) %>% 
  filter(n() >= 2) %>%      # require >=2 replicates per source
  summarise(
    d13C_mean = mean(d13C),
    d13C_sd   = sd(d13C),
    d15N_mean = mean(d15N),
    d15N_sd   = sd(d15N)
  ) %>% 
  ungroup()

# 5. Run simmr for each cat ---------------------------------------

# the mixtures (consumers) need to be a two column matrix
mix_cats <- cbind(cats$d13C, cats$d15N)

# the groups need to be integer codes that identify each group of consumers
# and correspond to the mix data above

####We are not comparing two different groups, so we do not need this 
#grp <- mix.data$Code

# the sources are split into a vector of names, and matrices of means
# and standard deviations
prey_prepped <- prey %>%
  mutate(name = "Prey",
         d13C = as.numeric(d13C),
         d15N = as.numeric(d15N)) %>% 
  select(name, d13C, d15N) %>% 
  group_by(name) %>% 
  summarise(name = "Prey",
    meand13C = mean(d13C),
            sdd13C = sd(d13C),
            meand15N = mean(d15N),
            sdd15N = sd(d15N))

wet_food <- food2 %>%
  mutate(type = case_when(type == "WetFish" ~ "Wet",
                          TRUE ~ type),
         name = case_when(type == "Wet" ~ "Wet",
                          TRUE ~ "Dry")) %>%
  filter(name == "Wet") %>% 
  group_by(name) %>% 
  summarise(name = "Wet",
            meand13C = mean(d13C),
            sdd13C = sd(d13C),
            meand15N = mean(d15N),
            sdd15N = sd(d15N))

dry_food <- food2 %>%
  mutate(type = case_when(type == "WetFish" ~ "Wet",
                          TRUE ~ type),
         name = case_when(type == "Wet" ~ "Wet",
                          TRUE ~ "Dry")) %>%
  filter(name == "Dry") %>% 
  group_by(name) %>% 
  summarise(name = "Dry",
            meand13C = mean(d13C),
            sdd13C = sd(d13C),
            meand15N = mean(d15N),
            sdd15N = sd(d15N))


cat_srcs <- rbind(prey_prepped, wet_food, dry_food) %>% 
  mutate(C_d13C_mean = 2.6,
         C_d15N_mean = 1.9,
         C_d13C_sd = 0,
         C_d15N_sd = 0)


s_names <- as.vector(cat_srcs$name)
s_means <- cbind(cat_srcs$meand13C, cat_srcs$meand15N)
s_sds   <- cbind(cat_srcs$sdd13C,   cat_srcs$sdd15N)

# the corrections are specified similarly but do not need a name vector
c_means <- cbind(cat_srcs$C_d13C_mean, cat_srcs$C_d15N_mean)
c_sds   <- cbind(cat_srcs$C_d13C_sd, cat_srcs$C_d15N_sd)

#McDonald et al did not include concentration means
#conc_means <- cbind(concs$meand13C, concs$meand15N)
#conc_sds   <- cbind(concs$sdd13C, concs$sdd15N)

# finally create our simmr object

#Checking whether these are True
#nrow(s_means) == length(s_names)  # must be TRUE
#ncol(s_means) == 2                # must be TRUE
#all(dim(s_means) == dim(s_sds))  # must be TRUE
#all(dim(s_means) == dim(c_means))

simmr_in = simmr_load(mixtures=mix_cats,
                     source_names=s_names,
                     source_means=s_means,
                     source_sds=s_sds,
                     correction_means=c_means,
                     correction_sds=c_sds)




plot(simmr_in)

simmr_out <- simmr_mcmc(simmr_in)

knitr::kable(summary(simmr_out))

# 1. Define your TEFs
tef_C <- 2.6   # ‰ for δ13C
tef_N <- 1.9   # ‰ for δ15N

# 2. Prepare consumer data (unchanged)
analysis_cat_data <- raw_cat_data %>% 
  filter(`Part of Analysis?` == "Y") %>% 
  mutate(
    d13C = as.numeric(d13C),
    d15N = as.numeric(d15N)
  )

library(FNN)


# Find 5 nearest neighbors for each observation
set.seed(123)  # for reproducibility
kmeans_result <- kmeans(analysis_cat_data[, c("d13C", "d15N")], centers = 2)

analysis_cat_clustered <- analysis_cat_data %>%
  mutate(cluster = as.factor(kmeans_result$cluster),
         cluster = case_when(cluster == 1 ~ "Hunter",
                             cluster == 2 ~ "Non Hunters"),
         paper_classification = case_when("Identifier" == "H84" ~ "",))

## Initial results
ggplot(analysis_cat_clustered, aes(x = d13C, y = d15N, color = cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    x = expression(delta^13*C),
    y = expression(delta^15*N),
    title = "Clusters of Cats by Isotope Values"
  ) +
  theme_minimal()

# 3. Prepare source data WITH TEF applied
analysis_food_data_tef <- raw_food_data %>% 
  filter(`Part of Analysis?` == "Y") %>% 
  mutate(
    d13C = as.numeric(d13C) + tef_C,
    d15N = as.numeric(d15N) + tef_N)

analysis_prey_data_tef <- raw_prey_data %>% 
  filter(`Part of Analysis?` == "Y") %>% 
  mutate(
    d13C = as.numeric(d13C) + tef_C,
    d15N = as.numeric(d15N) + tef_N
  )



ggplot() +
  # Cats: big blue dots
  geom_point(
    data = analysis_cat_data,
    aes(x = d13C, y = d15N, color = "Cats", shape = "Cats"),
    size  = 3,
    alpha = 0.8
  ) +
  # Pet food: small red triangles, more transparent
  geom_point(
    data = analysis_food_data_tef,
    aes(x = d13C, y = d15N, color = "Pet food", shape = "Pet food"),
    size  = 2.5,
    alpha = 0.5
  ) +
  # Prey: small green squares, more transparent
  geom_point(
    data = analysis_prey_data_tef,
    aes(x = d13C, y = d15N, color = "Prey", shape = "Prey"),
    size  = 2.5,
    alpha = 0.5
  ) +
  # Define the exact colors and shapes for each label
  scale_color_manual(
    name   = NULL,
    values = c("Cats" = "blue", "Pet food" = "red", "Prey" = "green")
  ) +
  scale_shape_manual(
    name   = NULL,
    values = c("Cats"     = 16,  # solid circle
               "Pet food" = 17,  # triangle
               "Prey"     = 15)  # square
  ) +
  labs(
    x     = expression(delta^13*C),
    y     = expression(delta^15*N),
    title = "Isotope Space: Consumers and TEF-Adjusted Sources"
  ) +
  theme_minimal() +
  lims(x = c(-21.3, -14),
       y = c(4.7, 10.2))



# Ensure cluster is a factor (for consistent coloring)
analysis_cat_clustered <- analysis_cat_clustered %>%
  mutate(cluster = as.factor(cluster))

library(dplyr)
library(ggplot2)

# Make sure cluster is a factor
analysis_cat_clustered <- analysis_cat_clustered %>%
  mutate(Source = case_when(
    cluster == 1 ~ "Hunter",
    cluster == 2 ~ "Non Hunter",
    ))

# Add a source column to food and prey
food_labeled <- analysis_food_data_tef %>%
  mutate(Source = "Pet food")

prey_labeled <- analysis_prey_data_tef %>%
  mutate(Source = "Prey")

combined_plot_data <- bind_rows(
  analysis_cat_clustered %>% select(d13C, d15N, Source, Name, Identifier),
  food_labeled %>% select(d13C, d15N, Source),
  prey_labeled %>% select(d13C, d15N, Source)
)

source_shapes <- c("Pet food" = 17, "Prey" = 15,
                   "Hunter" = 16, "Non Hunter" = 16)

source_colors <- c("Pet food" = "red", "Prey" = "green",
                   "Hunter" = "orange", "Non Hunter" = "purple")

p <- ggplot() +
  # Cats (Hunter & Non Hunter): larger, more opaque
  geom_point(
    data = combined_plot_data %>% filter(Source %in% c("Hunter", "Non Hunter")),
    aes(x = d13C, y = d15N, color = Source, shape = Source),
    size  = 3.5,
    alpha = 0.8
  ) +
  # Pet food: smaller, more transparent
  geom_point(
    data = combined_plot_data %>% filter(Source == "Pet food"),
    aes(x = d13C, y = d15N, color = Source, shape = Source),
    size  = 2,
    alpha = 6
  ) +
  # Prey: smaller, more transparent
  geom_point(
    data = combined_plot_data %>% filter(Source == "Prey"),
    aes(x = d13C, y = d15N, color = Source, shape = Source),
    size  = 2,
    alpha = 0.6
  ) +
  # 2. Add ellipses around the two cat groups
  stat_ellipse(
    data = combined_plot_data %>% filter(Source %in% c("Hunter", "Non Hunter")),
    aes(x = d13C, y = d15N, color = Source),
    type     = "norm",
    level    = 0.95,
    linetype = "dashed",
    size     = 1,
    inherit.aes = FALSE
  ) +
  # 3. Scales & labels
  scale_color_manual(values = source_colors) +
  scale_shape_manual(values = source_shapes) +
  labs(
    x     = expression(delta^13*C),
    y     = expression(delta^15*N),
    title = "Isotope Space: Hunter Status and TEF-Adjusted Sources",
    color = "Group",
    shape = "Group"
  ) +
  theme_minimal() +
  lims(x = c(-21.3, -14), y = c(4.7, 10.2))

# 4. Print
print(p)


combined_plot_data <- combined_plot_data %>%
  mutate(hover_text = paste0("Identifier: ", Identifier, "<br>Name: ", Name))


p <- ggplot() +
  geom_point(
    data = combined_plot_data %>% filter(Source %in% c("Hunter", "Non Hunter")),
    aes(
      x = d13C,
      y = d15N,
      color = Source,
      shape = Source,
      text = hover_text
    ),
    size  = 3.5,
    alpha = 0.8
  ) +
  labs(x = "δ13C", y = "δ15N") +
  lims(x = c(-21,-13), y = c(4,9))# Use Unicode symbols


# Convert to plotly with custom tooltip
ggplotly(p, tooltip = "text")

library(plotly)

table(analysis_cat_clustered$cluster)

datatable(posterior_probabilities)

```
McDonald Et al don't specifically mention how they calculated the Trophic Enrichment/Discriminatory Factor, but I am assuming they used this equation, (Fry, 2006): Δ=δc−δf where c refers to the isotopic signature/mixture of the consumer, and the f refers to the isotopic signature/mixture of the food source at equilibrium. Essentially, that occurs when the consumer has fed on the food long enough to where there signature values are no longer changing.


### Brian's Video Series

```{r, echo = FALSE, warning = FALSE, message = FALSE}

fish_example <- read_csv("prelim.csv")

fish_summary <- ddply(fish_example, c("Lake","Species"), summarise,
                      count = length(Species),
                      mc = mean(d13C), sdC = sd(d13C),
                      mN = mean(d15N), sdN = sd(d15N))

lake_summary <- ddply(fish_example, c("Lake"), summarise,
                      count = length(Species),
                      mc = mean(d13C), sdC = sd(d13C),
                      mN = mean(d15N), sdN = sd(d15N))

perch_Lsr <- fish_example %>% 
  filter(Species %in% c("lsr", "perch"))

ggplot(data = perch_Lsr, aes(x = Species, y = d13C, group = Species)) +
  geom_boxplot()+
  geom_jitter() +
  facet_grid(~ Lake)





```




```{r, echo = FALSE, warning = FALSE, message = FALSE}

data("geese_data")
mixtures_df <- as.data.frame(geese_data["mixtures"])
df_1 <- as.data.frame(geese_data["source_names"])
source_df <- as.data.frame(geese_data$source_means)
colnames(source_df) <- c("d13C", "d15N")
rownames(source_df) <- geese_data$source_names

source_df

simmr_in <- with(geese_data,
  simmr_load(
    mixtures = mixtures,
    source_names = source_names,
    source_means = source_means,
    source_sds = source_sds,
    correction_means = correction_means,
    correction_sds = correction_sds,
    concentration_means = concentration_means
  )
)




library(ggplot2)

# Mixtures (consumers)
mix_df <- as.data.frame(geese_data$mixtures)
colnames(mix_df) <- c("d13C", "d15N")

# Sources
src_df <- as.data.frame(geese_data$source_means)
colnames(src_df) <- c("d13C", "d15N")
src_df$Source <- geese_data$source_names

# Plot
ggplot() +
  geom_point(data = mix_df, aes(x = d13C, y = d15N), color = "blue", alpha = 0.6) +
  geom_point(data = src_df, aes(x = d13C, y = d15N), color = "red", shape = 17, size = 3) +
  geom_text(data = src_df, aes(x = d13C, y = d15N, label = Source), vjust = -1.2) +
  labs(x = expression(delta^13*C), y = expression(delta^15*N),
       title = "Isotope Space: Consumers and Sources") +
  theme_minimal()


plot(simmr_in) 

```


```{r, echo = FALSE, warning = FALSE, message = FALSE}

simmr_out <- simmr_mcmc(simmr_in)

summary(simmr_out)
```

### Cat Example Data

```{r, echo = FALSE, warning = FALSE, message=FALSE}

raw_cat_data <- read_csv("raw_cat_data.csv")
raw_food_data <- read_csv("raw_food_data.csv")
raw_prey_data <- read_csv("raw_prey_data.csv") %>% 
  mutate(d13C = as.numeric(d13C),
         d15N = as.numeric(d15N))

analysis_cat_data <- raw_cat_data %>% 
  filter(`Part of Analysis?` == "Y") %>% 
  mutate(d13C = as.numeric(d13C),
         d15N = as.numeric(d15N))

analysis_food_data <- raw_food_data %>% 
  filter(`Part of Analysis?` == "Y") %>% 
  mutate(d13C = as.numeric(d13C),
         d15N = as.numeric(d15N))

analysis_prey_data <- raw_prey_data %>% 
  filter(`Part of Analysis?` == "Y") %>% 
  mutate(d13C = as.numeric(d13C),
         d15N = as.numeric(d15N))

# Plot
ggplot() +
  geom_point(data = raw_cat_data, aes(x = d13C, y = d15N), color = "blue", alpha = 0.6) +
  geom_point(data = raw_food_data, aes(x = d13C, y = d15N), color = "red", shape = 17, size = 3) +
  geom_point(data = raw_prey_data, aes(x = d13C, y = d15N), color = "green", shape = 15, seize = 3) +
  #geom_text(data = src_df, aes(x = d13C, y = d15N, label = Source), vjust = -1.2) +
  labs(x = expression(delta^13*C), y = expression(delta^15*N),
       title = "Isotope Space: Consumers and Sources") +
  theme_minimal()


# Plot
ggplot() +
  geom_point(data = analysis_cat_data, aes(x = d13C, y = d15N), color = "blue", alpha = 0.6) +
  geom_point(data = analysis_food_data, aes(x = d13C, y = d15N), color = "red", shape = 17, size = 3) +
  geom_point(data = analysis_prey_data, aes(x = d13C, y = d15N), color = "green", shape = 15, seize = 3) +
  #geom_text(data = src_df, aes(x = d13C, y = d15N, label = Source), vjust = -1.2) +
  labs(x = expression(delta^13*C), y = expression(delta^15*N),
       title = "Isotope Space: Consumers and Sources") +
  theme_minimal()



# 1. Define your TEFs
tef_C <- 2.6   # ‰ for δ13C
tef_N <- 1.9   # ‰ for δ15N

# 2. Prepare consumer data (unchanged)
analysis_cat_data <- raw_cat_data %>% 
  filter(`Part of Analysis?` == "Y") %>% 
  mutate(
    d13C = as.numeric(d13C),
    d15N = as.numeric(d15N)
  )

# 3. Prepare source data WITH TEF applied
analysis_food_data_tef <- raw_food_data %>% 
  filter(`Part of Analysis?` == "Y") %>% 
  mutate(
    d13C = as.numeric(d13C) + tef_C,
    d15N = as.numeric(d15N) + tef_N
  )

analysis_prey_data_tef <- raw_prey_data %>% 
  filter(`Part of Analysis?` == "Y") %>% 
  mutate(
    d13C = as.numeric(d13C) + tef_C,
    d15N = as.numeric(d15N) + tef_N
  )



ggplot() +
  # Cats: big blue dots
  geom_point(
    data = analysis_cat_data,
    aes(x = d13C, y = d15N, color = "Cats", shape = "Cats"),
    size  = 3,
    alpha = 0.8
  ) +
  # Pet food: small red triangles, more transparent
  geom_point(
    data = analysis_food_data_tef,
    aes(x = d13C, y = d15N, color = "Pet food", shape = "Pet food"),
    size  = 2.5,
    alpha = 0.5
  ) +
  # Prey: small green squares, more transparent
  geom_point(
    data = analysis_prey_data_tef,
    aes(x = d13C, y = d15N, color = "Prey", shape = "Prey"),
    size  = 2.5,
    alpha = 0.5
  ) +
  # Define the exact colors and shapes for each label
  scale_color_manual(
    name   = NULL,
    values = c("Cats" = "blue", "Pet food" = "red", "Prey" = "green")
  ) +
  scale_shape_manual(
    name   = NULL,
    values = c("Cats"     = 16,  # solid circle
               "Pet food" = 17,  # triangle
               "Prey"     = 15)  # square
  ) +
  labs(
    x     = expression(delta^13*C),
    y     = expression(delta^15*N),
    title = "Isotope Space: Consumers and TEF-Adjusted Sources"
  ) +
  theme_minimal()



```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

# source$data_type: means
# source$by_factor: NA
# random effects: 0
# fixed effects: 0
# nested factors: 
# factors: 
# continuous effects: 0
# error structure: Process only (MixSIR, for N = 1)
# source$conc_dep: FALSE



# model{
#   for(src in 1:n.sources){
#     for(iso in 1:n.iso){
#       src_mu[src,iso] ~ dnorm(MU_array[src,iso], n_array[src]/SIG2_array[src,iso]);  # Eqn 3.8 but with precision instead of variance
#       tmp.X[src,iso] ~ dchisqr(n_array[src]);
#       src_tau[src,iso] <- tmp.X[src,iso]/(SIG2_array[src,iso]*(n_array[src] - 1));   # Eqn 3.9, following the simulation on p.580
#     } 
#   }
# 
#     # draw p.global (global proportion means) from an uninformative Dirichlet,
#     # then ilr.global is the ILR-transform of p.global
#     p.global[1:n.sources] ~ ddirch(alpha[1:n.sources]);
#     for(src in 1:(n.sources-1)){
#       gmean[src] <- prod(p.global[1:src])^(1/src);
#       ilr.global[src] <- sqrt(src/(src+1))*log(gmean[src]/p.global[src+1]); # page 296, Egozcue 2003
#     }
# 
#    # DON'T generate individual deviates from the global/region/pack mean (but keep same model structure)
#    for(i in 1:N) {
#       for(src in 1:(n.sources-1)) {
#          ilr.ind[i,src] <- 0;
#          ilr.tot[i,src] <- ilr.global[src] + ilr.ind[i,src]; # add all effects together for each individual (in ilr-space)
#       }
#    }
# 
#    # Inverse ILR math (equation 24, page 294, Egozcue 2003)
#    for(i in 1:N){
#       for(j in 1:(n.sources-1)){
#         cross[i,,j] <- (e[,j]^ilr.tot[i,j])/sum(e[,j]^ilr.tot[i,j]);
#       }
#       for(src in 1:n.sources){
#         tmp.p[i,src] <- prod(cross[i,src,]);
#       }
#       for(src in 1:n.sources){
#         p.ind[i,src] <- tmp.p[i,src]/sum(tmp.p[i,]);
#       }
#    }
# 
#    for(src in 1:n.sources) {
#       for(i in 1:N){
#          # these are weights for variances
#          p2[i,src] <- p.ind[i,src]*p.ind[i,src];
#       }
#    }
# 
# 
#    # for each isotope and population, calculate the predicted mixtures
#    for(iso in 1:n.iso) {
#       for(i in 1:N) {
# 
#          mix.mu[iso,i] <- inprod(src_mu[,iso],p.ind[i,]) + inprod(frac_mu[,iso],p.ind[i,]);
#       }
#    }
# 
# 
#   # calculate mix variance and likelihood
#   for(i in 1:N){
#     for(iso in 1:n.iso){
# 
#       process.var[iso,i] <- inprod(1/src_tau[,iso],p2[i,]) + inprod(frac_sig2[,iso],p2[i,]);
#       mix.prcsn[iso,i] <- 1/process.var[iso,i];
#       X_iso[i,iso] ~ dnorm(mix.mu[iso,i], mix.prcsn[iso,i]);
#       loglik_mat[i,iso] <- logdensity.norm(X_iso[i,iso], mix.mu[iso,i], mix.prcsn[iso,i]);
#     }
#     loglik[i] <- sum(loglik_mat[i,])
#   }
# }  # end model


```

